If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-08 GroupPolarizationStatmod/paper/ms.tex
updates; continuing draft

\usepackage{authblk}
\usepackage{etoolbox}
\usepackage{wrapfig}
\makeatletter
\title{Group polarization replications may be marred by high false discovery rates}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Division of Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
\noindent 
\emph{Group polarization} is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important because
it may increase social and political tensions if moderates become more
extreme. Decades of studies have replicated detections of group
polarization, typically using an ordinal scale to measure opinions.
However, these studies did not account for ordinal measurements, which can
result in spurious detections of opinion change.  Lacking original data we
can still calculate the probability that a published detection of group
polarization was spurious, i.e., to calculate the \emph{false detection
rate}.  Across 54 group polarization experimental conditions from ten
representative journal articles, we calculated false detection rates to be
between 0.52 and 0.88, with a median of 0.75, using a generative model of
group polarization experiments seeded with empirical data.  
We also use our model to develop experimental designs that
achieve an acceptable false discovery rate.  Much of group polarization
research may be unreliable. This analysis can help change that by enabling others to avoid this
hidden pitfall. More
broadly this work demonstrates one important way that replication success alone
does not imply epistemic reliability.
\end{abstract}

@@

\begin{center} \noindent\rule{4cm}{0.4pt} \end{center}
\clearpage

@@

\section{Introduction}\label{introduction}

If an extremist's opinion falls and it is measured with a Likert scale, will it
make a noise? The answer had better be yes if we want to properly measure opinion
change. Accurate measurements are critical
for developing rigorous and reliable methods to promote more centrist,
sustainable, and responsive governments and institutions
\autocite{Mason2018UncivilAgreementBook,Klein2020}. Unfortunately, social and
behavioral scientists have often used statistical methods 
that distort
opinion change when it's measured on an ordinal, Likert-style scale (e.g. -3
indicates ``Strongly disagree'', +3 indicates ``Strongly agree'', and 0 is
neutral)~\autocite{Liddell2018}. 
This includes research that Obama official and law professor Cass Sunstein
marshals to explain why ``people become extremists'' and why ``political and
cultural polarization'' is ``so pervasive in America''\footnote{From the
  Publisher's Summary. Sunstein does not make such general claims in the book,
  but does suggest that the research he reviews collectively provides
  a ``clue'' to explaining how
  ``facism...student radicalism...Islamic terrorism...[t]he Rwandan
genocide...[e]thnic Conflict.}
administration official and law professor Cass Sunstein writing a book called
*researchers construct

\emph{Group polarization} is that essentially
measures how much being in an opinion echo-chamber the name given the social phenomenon of 
when a group becomes more extreme in mean opinion over
time, assuming the mean opinion is biased to one extreme to begin with.
\mt{now hit them with the real shit to intro the problem, our analysis, and
how this can be used constructively going forward.}

\mt{Seems like a good second paragraph structurally. We should give more
details about the scope of the study, group polarization and common
elements of experimental designs.} 
Group polarization experiments tend to employ a common set of design
elements incorporated in our computational model analysis. Group
polarization experiments mostly use ordinal (see exceptions
\cite{Blascovich1974,Blascovich1975,Blascovich1976,Keating2016}). The opinions (e.g., ``Do you approve,
disapprove, or neither approve nor disapprove of the current
administration's policies?''). Extremity only increases, though, assuming
that the mean opinion of the group is initially biased towards one extreme
or the other \autocite{Brown1986,Brown2020}. Group polarization is important
to understand rigorously, therefore, since if groups become increasingly
extreme over time, while group opinions are consolidated, this would
exacerbate political polarization between political opposition groups, which
is already undermines the ability of partisans in democratic governments to
work together. Dozens of studies and meta-analyses over decades have
apparently established the empirical reality of group polarization
\autocite{Moscovici1969,Myers1970,Isenberg1986,Sunstein2009,Sunstein2019}
\autocite[Ch. 5]{Brown2000}. We believe group polarization should occur in
certain contexts since more extreme group members are likely also to be more
stubborn \autocite{Guazzini2015,Lewandowsky2019}, which should pull
consensus to extremes \autocite{Turner2020}. However, many or most of the
evidence for group polarization relies on experimental designs that use
ordinal measurements that are then analyzed as if they were continuous.
Unfortunately, this procedure is unfortunately known to induce false
positives if it is not properly accounted for \autocite{Liddell2018}.

The null-consistency procedure functions as a test of model indistinguishability: it demonstrates that a reported group polarization effect could plausibly arise from a stable-latent process combined with changing opinion precision. This logic echoes foundational concerns in causal inference, where two competing models—one causal, one not—can imply the same observed data distribution and are therefore empirically indistinguishable without further assumptions \cite{pearl2000causality,spirtes2000causation}. In such cases, inference from observation to mechanism is invalid. Similarly, in psychometrics, it is well known that different configurations of latent traits and response thresholds can yield identical observed ordinal patterns, making the true structure unidentifiable without additional constraints \cite{borsboom2005measuring}. The present approach does not attempt to estimate the probability of false detection, but rather shows that under ordinal measurement, a null model can reproduce the observed pattern—rendering the result non-diagnostic of a true latent shift. This establishes a stringent standard: an effect must not only be detected, but shown to be incompatible with plausible null-generating mechanisms.

Experimental designs in group polarization studies share a certain
experimental boilerplate (Figure 1 --- NEED TO INCLUDE) with the goal of
testing the group polarization hypothesis that a group of like-minded
people will develop more extreme opinions on some salient topic over the
course of discussing some topic (Figure~\ref{fig:ModelSketch}). 
Group polarization experiments are thus
initialized by forming like-minded groups of people and surveying their
initial opinions on some topic. In some cases these initial opinions are
used for assigning participants to like-minded groups, but other methods
have been used, such as geographic location for producing, on average,
like-minded groups \autocite{Schkade2010}. Participants then discuss the
topic in their groups. After discussion, participant opinions are again
measured. Several studies we analyze here determined group polarization
occurred (i.e., defined a positive detection) only as a non-zero
difference in mean opinion, with no other statistical testing. Some
others performed null-hypothesis \(t\)-tests to test whether there was
enough evidence to reject the null-hypothesis that \emph{simple
consensus} occurredand not group polarization. In our meta-analysis
simulations, we measure simulated group polarization effect size in
terms of Cohen's \(d\), which is an abstraction beyond these two
practices from the group polarization literature that can also be used
to compare model fits between metric models that treat group
polarization measurements as continuous although they are ordinal, and
more appropriate statistical models that account for data ordinality.

% \begin{figure}
%   \caption{\textbf{Opinions are implicitly behavioral choices that are most consistent with some
%   complex aggregation of brain states and dynamics, and so are theoretically
%   continuous (A). Common components of a group polarization experiment.}
%   Participants first give their opinion in response to some prompt (B). Afterwards,
%   participants deliberate the topic (C). Participants then give their opinion again
%   (D). }
%   \centering
%     \includegraphics[width=0.65\textwidth]{Figures/Model/ModelSketchesA.pdf}
%     \includegraphics[width=0.7\textwidth]{Figures/Model/ModelSketchesB-D.pdf}
%   \label{fig:ModelSketch}
% \end{figure}

In this paper, we perform a meta-analysis of sixty detections of group
polarization from ten published research articles and show that
fifty-five of the sixty cannot be reliably identified as group
polarization. We show that these 55 are plausibly false detections
because they are equally well-explained by a \emph{simple consensus}
model, where the mean group opinion is unchanged by group discussion but
opinion variance decreases. Theoretically, consensus occurs when
sufficiently like-minded individuals influence, which decreases group
opinion variance \autocite{DeGroot1974,Turner2018}. Group polarization,
then, is consensus plus an overall increase in group opinion extremity,
i.e., an increase in mean group opinion after discussion. We use a novel
approach, described herein, that enabled us to identify continuous pre-
and post-discussion simple consensus opinion distributions that result
in data that appear to be group-polarized when opinions are drawn from
these distributions and measured with an ordinal scale. This approach
could be adapted for similar social psychological studies that used
ordinal measures to determine whether change occurred in distributions
of continuous psychological variables such as opinions and beliefs. For
social psychology and group polarization, this means that an accepted,
heretofore unquestioned, empirical fact must be re-evaluated.
Experimental designs in the field must be updated if we are to
rigorously understand whether and when group polarization occurs.

\subsection{Studies included in this
meta-analysis}\label{studies-included-in-this-meta-analysis}

We chose the ten studies guided by a desire to identify foundational
studies that contribute to the general consensus in social psychology
that group polarization reliably occurs across experimental
settings---one author of one of the ten journal articles is a law
professor who has written two trade books on group polarization to argue
that group polarization occurs in real-world settings in addition to the
laboratory, with political and other practical importance
\autocite{Sunstein2009,Sunstein2019}. We chose studies that either had a
significant influence on the study of group polarization and social
psychology, in terms of a high citation count or adaptation in
popularized trade books, or were designed to support one of four
prominent theoretical explanations of group polarization. The studies
must also have used an ordinal scale for measuring participant opinions
(all but a handful of exceptions use ordinal measurements to measure
group polarization; e.g.~blackjack bets
\autocite{Blascovich1973,Blascovich1976} and mock jury monetary damages
\autocite{Schkade2000}). We included at least one study that supported
or was motivated by one of four established theoretical explanations
(introduced below) of group polarization to show all have at least some
supporting evidence that must be discarded. We also included two studies
that lack a clear theoretical motivation. For the sake of orienting the
reader to the group polarization literature and as a useful organizing
feature for our analysis, we now introduce the theories that motivate
eight of the ten studies. To further orient the reader, we then provide
details of the experimental designs used to produce the 60 published
positive detections of group polarization across 10 research articles
that we scrutinize here. Group polarization experimental designs are
diverse and inconsistent with one another in many ways, which
complicates this or other potential meta-analyses and may imperil our
ability to rigorously understand how regular the group polarization
effect really is \autocite{Yarkoni2022,Almaatouq2022}.

The first theoretical explanation whose supporting findings we include
here is (1) social comparisons theory, which posits that an individual's
opinions are updated so they stand out in a group enough to be
``desirably distinctive'' (Myers, 1978, p.~562) but not too distinct to
be viewed as anti-conformist \autocite{Brown1974,Sanders1977,Myers1978}.
Second, \emph{persuasive arguments theory} posits that more extreme
individuals have a greater number of more convincing arguments for their
view, which draw more moderate group members to extreme opinions
\autocite{Burnstein1973,Burnstein1975,Burnstein1977}. Third, the
\emph{self-categorization theory} predicts that group polarization will
occur as more moderate individuals judge the mean group opinion to be
more extreme \autocite{Turner1989,Abrams1990,Krizan2007}. Fourth and
finally, the social decision schemes explanation posits that the
structure of \autocite{Davis1973,Zuber1992,Friedkin1999a}. We do not
evaluate or compare these theories here, though a simpler explanation
that extremists are more stubborn may be a more parsimonious explanation
consistent with these four alternatives
\autocite{Acemoglu2013,Guazzini2015,Turner2020}.

Each of the ten group polarization studies we analyze here varied
significantly in how groups were formed, how opinions were gathered, and
other details. There was significant variation across experimental
designs, participant population sizes, group sizes, and how like-minded
groups were formed in the first place. We review some of these details
below to provide the reader with some insight into the experimental
designs that produced what we show are plausibly false positive
detections of group polarization. We focus here on the questions that
were asked, how like-minded novel experimental groups were formed,
organizing our presentation by theoretical tradition.

\subsubsection{Details of experimental conditions included in this
meta-analysis}\label{details-of-experimental-conditions-included-in-this-meta-analysis}

\paragraph*{Atheoretical}

The first of ten studies we included in our meta-analysis was by
Moscovici and Zavalloni (1969) \autocite{Moscovici1969} who sought to
measure group polarization among groups of French high school students
who discussed domestic and foreign politics. Moscovici and Zavalloni
tested four conditions whether group polarization occurred for two
different questions and two different rating systems: participants were
asked their opinions on then-president Charles De Gaulle and their
opinion about foreign aid from the USA either in the ``Opinion'' setting
or the ``Judgement'' condition (see Table 4 on p.~132).

The second, relatively more recent article is Schkade, Sunstein, and
Hastie (2010) \autocite{Schkade2010} who used geographic location to
construct like-minded groups by running a group polarization experiment
in two cities in the state of Colorado in the United States: Colorado
Springs, whose residents historically voted for republican candidates,
and Boulder, whose residents historically voted for democratic
candidates. Discussion groups were formed in which they measured group
polarization when group members discussed global warming, affirmative
action, and same-sex civil unions.

\paragraph{Social comparisons}

In developing the social comparisons theory of group polarization
(described in Myers (1978) \autocite{Myers1978}), Myers and Bishop
(1970), writing in \emph{Science,} \autocite{Myers1970} sought to
measure group polarization among high school students in the United
States who discussed racial attitudes and Myers (1975)
\autocite{Myers1975} measured group polarization in groups of college
students discussing the quality of a faculty member in one condition,
and women's rights and issues in another condition.

\paragraph{Persuasive arguments}

\paragraph{Self-categorization}

\paragraph{Social decisions scheme}

The ten studies we included each tested several experimental conditions
meant to elicit group polarization in novel experimental groups, 60 in
total, which we briefly introduce here.\\

\subsection{Research overview}\label{research-overview}

To evaluate whether published detections of group polarization are
plausibly false we developed and combined formal and generative models
of group polarization to identify plausible simple consensus pre- and
post-discussion opinion distributions that seem like group polarization
when binned or measured on ordinal scales. Opinion distributions are
assumed to be normal. We define a \emph{simple consensus pair} of
\emph{latent opinion} distributions to be a pair of pre- and
post-discussion distributions where the mean opinion is constant, but
the pre-discussion variance is greater than the post-discussion
variance. A \emph{latent distribution} of opinions represents the
internal, unobservable distribution of opinions for a group or
collection of groups. The \emph{observed distribution} is the
distribution of reported ordinal opinion measurements. We then define a
\emph{plausibly false detection} of group polarization to be one where
the mean post-discussion observed distribution is greater in magnitude
(i.e., more extreme) than the post-discussion observed distribution's,
but both were measurements of a pair of latent simple consensus
distributions.

In the formal model, a simulated observed distribution is generated by
binning a latent distribution into a finite number of ordinal scale
bins, then integrating over the probability density in each bin. In the
generative model, an observed distribution is simulated by first drawing
participant opinions from a latent distribution, then counting the
number of opinions falling within each bin. The formal model is used to
identify which, if any, simple consensus pairs exist for a given
reported detection of group polarization through an optimization routine
that finds pre- and post-discussion variances that generate simulated
observations that appear to be group polarization (or reports a failure
to such variances). If the optimization routine finds a constant latent
mean that generates simulated observed group polarization, then we say
the published group polarization detection is plausibly false. If the
routine fails to find a constant latent mean that generates simulated
group polarization, then we cannot and do not claim the detection is
plausibly false. We use the generative model to evaluate how frequently
we would expect plausibly false detections when there are finite-size
groups, since the formal model implicitly applies infinite-size groups.
Therefore in our analyses, we also generate one thousand simulated group
polarization datasets for each of the 55 plausibly false detections
identified with the formal model and examine the distribution of group
polarization effect sizes in terms of Cohen's \(d\). In the first round
of simulations we set the number of observations to match those in the
ten published studies we analyzed. Many or most simulated effect sizes
are non-zero with these numbers of observations. To see if perhaps these
non-zero effects were due to low statistical power from too few
observations, we also tested a hypothetical study with ten times more
participants. Finally, we evaluate whether an appropriate statistical
model that accounts for the ordinal measurement of latent opinions can
reliably identify simulated simple consensus in latent opinions when
simulated observations appear as group polarization based on mean group
opinion alone.

% \begin{wrapfigure}{l}{0.5\textwidth}
\begin{figure}
  \caption{\textbf{Group polarization experimental model.} Latent opinions are 
    represented as numbers corresponding to the valence and extremism of one's opinion
(A). Before deliberating a topic of interest, all participants report their 
opinion (B). Participants then deliberate the topic with others (C). Following
deliberation participants report their opinion once more (D).}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/Model/ModelSketches.pdf}
  \label{fig:experiment_model}
% \end{wrapfigure}
\end{figure}

>>>>

\usepackage{authblk}
\usepackage{etoolbox}
\usepackage{wrapfig}
\makeatletter
\title{If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization}
\makeatother
\author[1,*]{{Matthew A.~Turner}}
\affil[1]{\small Environmental Social Sciences, Stanford Doerr School of Sustainability, Stanford University}

@@

\date{}
\begin{document}
\maketitle
% \begin{abstract}
% \noindent 
% \emph{Group polarization} is the name for a form of consensus where
% members of a like-minded group become, on average, more extreme in their
% opinions after discussing a topic. Group polarization is important because
% it may increase social and political tensions if moderates become more
% extreme. Decades of studies have replicated detections of group
% polarization, typically using an ordinal scale to measure opinions.
% However, these studies did not account for ordinal measurements, which can
% result in spurious detections of opinion change.  Lacking original data we
% can still calculate the probability that a published detection of group
% polarization was spurious, i.e., to calculate the \emph{false detection
% rate}.  Across 54 group polarization experimental conditions from ten
% representative journal articles, we calculated false detection rates to be
% between 0.52 and 0.88, with a median of 0.75, using a generative model of
% group polarization experiments seeded with empirical data.  
% We also use our model to develop experimental designs that
% achieve an acceptable false discovery rate.  Much of group polarization
% research may be unreliable. This analysis can help change that by enabling others to avoid this
% hidden pitfall. More
% broadly this work demonstrates one important way that replication success alone
% does not imply epistemic reliability.
% \end{abstract}

@@

\begin{center} \noindent\rule{4cm}{0.4pt} \end{center}
% \clearpage

@@

\input{intro-latest.tex}

% \input{old-intro.tex}
